---
title: Skagit Steelhead Test Fishery Data Exploration
author: Casey Ruff (cruff@swinomish.nsn.us) & Thomas Buehrens (tbuehrens@dfw.wa.gov)
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***


# Overview
This script fits Generalized Additive Models to Skagit Steelhead Test Fishery Data with and without the Total Run Size as an offset. The best model without TRS as an offset includes a spline term for year, which attempts to explain the residual variance (e.g., due to interannual run size differences) when the TRS offset is removed. Therefore, this spline term may be considered somewhat analogous to the model's ability to predict runsize based on the in-season test fishery data.

The code repository used to generate this page and complete analyses can be found here: [**(link)**](https://github.com/casruff/test_fishery_data_exploration)

# Setup
All analyses require R software [**(link)**](https://cran.r-project.org/) (v3.4.3) for data retrieval, data processing, and summarizing model results. Here we configure R to perform our analysis and generate our outputs
```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```

We also need a couple of helper functions which we will define
```{r load_funcs, message = FALSE, warning = FALSE,results = "hide"}
#function to install or load packages
install_or_load_pack <- function(pack){
  create.pkg <- pack[!(pack %in% installed.packages()[, "Package"])]
  if (length(create.pkg))
    install.packages(create.pkg, dependencies = TRUE)
  sapply(pack, require, character.only = TRUE)
}

```

Here we will load & install packages we need to use (needs internet connection if packages not already installed)
```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
packages_list<-c("tidyverse"
                 ,"forecast"
                 ,"mgcv"
                 ,"ggplot2"
                 ,"MASS"
                 ,"RColorBrewer"
                 ,"kableExtra"
                 ,"gtools"
                 ,"readr"
                 ,"here"
                 ,"readxl"
                 ,"brms"
                 ,"bsplus"
                 ,"lubridate"
                 ,"MuMIn"
                 ,"tidymv"
                 # ,"rnoaa"
                 # ,"ncdf4"
                 # ,"ncdf4.helpers"
                 # ,"raster"
                 # ,"reshape2"
                 # ,"ggfortify"
                 )
install_or_load_pack(pack = packages_list)
```

# User Inputs
```{r user_inputs, message = FALSE, warning = FALSE,results = "show"}
## set data dir
datadir <- here("data")

## first & last years of fish data
yr_frst <- 2013
yr_last <- 2020


## data file names
## 1. file with escapement data
fn_trs <- "sthd_trs.csv"

## 2. file with age comp data
fn_tst_fsh <- "skgt_sthd_tst_fsh.csv"

## get flow data (currently using mainstem at mount vernon)
## flow gage ID
flow_site <- 12200500  
```

# Load Data
```{r load_data, message = FALSE, warning = FALSE,results = "show"}
## read total runsize data
dat_trs <- read_csv(file.path(datadir,fn_trs))
## read in test fishery data
dat_tst_fsh <- read_csv(file.path(datadir,fn_tst_fsh))

## get URL for flow data from USGS
flow_url <- paste0("https://waterdata.usgs.gov/nwis/dv",
                   "?cb_00060=on",
                   "&format=rdb",
                   "&site_no=",flow_site,
                   "&begin_date=",yr_frst,"-01-01",
                   "&end_date=",yr_last,"-12-31")

## raw flow data from USGS
skip <- read_lines(flow_url)%>%
  as_tibble()%>%
  filter(grepl("\\#",value))%>%
  summarise(skip=n())%>%
  unlist()

## flow data for years of interest
dat_flow <-  read_tsv(flow_url,
                      col_names = FALSE,
                      col_types = "ciDdc",
                      skip = skip + 2 )%>%
  dplyr::rename(date=X3,flow=X4)%>%
  dplyr::select(date,flow)%>%
  mutate(flow = flow / 35.3147,
         year= year(date),
         day = yday(date),
         month = month(date)
  )

```

# Analysis by set (only using set 1 & 2)
```{r Analysis, message = FALSE, warning = FALSE,results = "show"}
dat<-dat_tst_fsh%>%
  left_join(dat_trs)%>%
  left_join(dat_flow)%>%
  mutate(site=factor(site),
         zl_flow=scale(log(flow)),
         effort = as.numeric(time_out-time_in)
         )%>%
  filter(set_num < 3 )#%>%
  #group_by(year,day,site,flow)%>%
  #summarise(SthdW = sum(SthdW),
  #          effort = sum(effort)      
  #          )
  

m1<-gam(SthdW ~  s(day) + s(site,bs="re"), offset = log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m2<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m3<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow) + s(set_num,bs="re"), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m4<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow) + s(set_num,bs="re")  + s(day,site,bs="fs"), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m5<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow) + s(set_num,bs="re")   + s(day,site,bs="fs") +  s(year,k=7,bs="ps",m=1), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)


model.sel(m1,m2,m3,m4,m5)


summary(m3)
plot(m3)
gam.check(m3)
plot(m3$y~ m3$fitted.values)
abline(a=0,b=1)

#best model removing offset of total run size, adding a year spline...this term should explain variance accounted for previously by offset term

m3b<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow) + s(set_num,bs="re") + s(year,k=7,bs="ps",m=1), offset = log(effort),  family = nb(theta = NULL, link = "log"),data=dat)


model.sel(m1,m2,m3,m4,m5,m3b)

#what does year spline term look like?
plot(m3b,select = 5,shade=T)
```

# Analysis by date/location (pooling sets)
```{r Analysis_v2, message = FALSE, warning = FALSE,results = "show"}
dat<-dat_tst_fsh%>%
  left_join(dat_trs)%>%
  left_join(dat_flow)%>%
  mutate(site=factor(site),
         zl_flow=scale(log(flow)),
         effort = as.numeric(time_out-time_in)
         )%>%
  filter(set_num < 7)%>%
  group_by(year,day,site,zl_flow,trs)%>%
  summarise(SthdW = sum(SthdW),
           effort = sum(effort)
           )
  

m1<-gam(SthdW ~  s(day) + s(site,bs="re"), offset = log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m2<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m3<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow) + s(day,site,bs="fs"), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)

m4<-gam(SthdW ~  s(day) + s(site,bs="re") + s(zl_flow) + s(day,site,bs="fs") +  s(year,k=7,bs="ps",m=1), offset=log(trs) + log(effort), family = nb(theta = NULL, link = "log"),data=dat)


model.sel(m1,m2,m3,m4)


summary(m1)
plot(m1)
gam.check(m1)
plot(m1$y~ m1$fitted.values)
abline(a=0,b=1)
#best model removing offset of total run size, adding a year spline...this term should explain variance accounted for previously by offset term

m1b<-gam(SthdW ~  s(day) + s(site,bs="re") + s(year,k=7,bs="ps",m=1), offset = log(effort),  family = nb(theta = NULL, link = "log"),data=dat)


model.sel(m1,m2,m3,m4,m1b)

#what does year spline term look like?
plot(m1b,select = 3,shade=T)
```

